---
title: "Transition_to_python"
author: "Amira Abozaid"
date: "`r Sys.Date()`"
output: html_document
---
```{r 00_set_root, message=FALSE, warning=FALSE}
setwd("E:/Breast_cancer_DL/METABRIC_DL_PAM50")
getwd()


```

---

## Chunk 01 — Minimal config + ONE export folder

```{r 01_config_and_one_export_folder, message=FALSE, warning=FALSE}

project <- list(
  dataset = "METABRIC_DL_PAM50",
  root    = normalizePath(getwd(), winslash = "/", mustWork = TRUE)
)

paths <- list(
  raw   = file.path(project$root, "01_raw"),
  out   = file.path(project$root, "EXPORT_FOR_COLAB")  # <— SINGLE OUTPUT FOLDER
)

dir.create(paths$out, recursive = TRUE, showWarnings = FALSE)
stopifnot(dir.exists(paths$raw), dir.exists(paths$out))

project
paths
list.files(paths$raw)


```

## Chunk 02 — Packages

```{r 02_packages, message=FALSE, warning=FALSE}

cran_pkgs <- c("data.table","dplyr","tibble","stringr","readr","tidyr","jsonlite","digest")
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)

invisible(lapply(cran_pkgs, library, character.only = TRUE))


```




## Chunk 03 — Locate / extract raw cBioPortal archive (self-healing)
This handles either:
- you have `brca_metabric.tar.gz` in `01_raw`, OR
- you already extracted into `01_raw/cbioportal_metabric/brca_metabric`

```{r 03_extract_or_use_existing, message=FALSE, warning=FALSE}
stopifnot(dir.exists(paths$raw))

extract_dir <- file.path(paths$raw, "cbioportal_metabric")
study_dir   <- file.path(extract_dir, "brca_metabric")

# If already extracted, keep it
if (!dir.exists(study_dir)) {

  # Find archive in 01_raw
  candidates <- list.files(
    paths$raw,
    pattern = "brca_metabric.*\\.(tar|tar\\.gz|tgz|gz)$",
    full.names = TRUE,
    ignore.case = TRUE
  )

  # Fallback: any tar-ish file
  if (length(candidates) == 0) {
    candidates <- list.files(paths$raw, pattern="\\.(tar|tar\\.gz|tgz|gz)$",
                             full.names=TRUE, ignore.case=TRUE)
  }

  stopifnot(length(candidates) >= 1)
  tar_fp <- candidates[which.max(file.info(candidates)$size)]

  dir.create(extract_dir, recursive = TRUE, showWarnings = FALSE)
  utils::untar(tar_fp, exdir = extract_dir)
}

stopifnot(dir.exists(study_dir))

expr_fp <- file.path(study_dir, "data_mrna_illumina_microarray.txt")
clp_fp  <- file.path(study_dir, "data_clinical_patient.txt")
cls_fp  <- file.path(study_dir, "data_clinical_sample.txt")

stopifnot(file.exists(expr_fp), file.exists(clp_fp), file.exists(cls_fp))

c(expr_fp=expr_fp, cl_patient=clp_fp, cl_sample=cls_fp)


```



## Chunk 04 — Robust clinical reader + load expression/clinical from raw

```{r 04_load_raw_expression_and_clinical, message=FALSE, warning=FALSE}

norm_id <- function(x) stringr::str_replace_all(as.character(x), "[^A-Za-z0-9_\\-]", "_")

pick_col <- function(df, candidates) {
  hit <- intersect(candidates, names(df))
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

pick_by_regex <- function(nms, patterns) {
  nms_low <- tolower(nms)
  for (p in patterns) {
    hit <- nms[nms_low %>% stringr::str_detect(p)]
    if (length(hit) > 0) return(hit[1])
  }
  NA_character_
}

read_cbio_clinical <- function(fp) {
  lines <- readLines(fp, warn = FALSE)
  lines <- lines[!is.na(lines)]
  lines_ne <- lines[nzchar(trimws(lines))]

  idx_data_start <- which(!stringr::str_detect(lines_ne, "^#"))[1]
  if (is.na(idx_data_start)) stop("No data lines found in: ", fp)

  tf <- tempfile(fileext = ".txt")
  writeLines(lines_ne, tf)
  dt <- data.table::fread(tf, skip = idx_data_start - 1, header = TRUE,
                          data.table = FALSE, fill = TRUE)
  unlink(tf)

  names(dt) <- make.names(trimws(names(dt)))
  dt
}

# ---- expression ----
expr_dt <- data.table::fread(expr_fp, data.table = FALSE)
colnames(expr_dt)[1] <- "gene"
expr_dt$gene <- toupper(trimws(as.character(expr_dt$gene)))

# coerce non-gene cols to numeric
non_gene_cols <- setdiff(names(expr_dt), "gene")
bad_cols <- non_gene_cols[!vapply(expr_dt[non_gene_cols], is.numeric, logical(1))]
if (length(bad_cols) > 0) {
  expr_dt[bad_cols] <- lapply(expr_dt[bad_cols], function(x) suppressWarnings(as.numeric(x)))
}

# collapse duplicate genes (median)
expr_dt <- expr_dt %>%
  dplyr::filter(!is.na(gene), gene != "") %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(dplyr::across(dplyr::where(is.numeric), ~ median(.x, na.rm = TRUE)), .groups = "drop")

expr_mat <- as.matrix(expr_dt[, setdiff(names(expr_dt), "gene"), drop = FALSE])
rownames(expr_mat) <- expr_dt$gene
storage.mode(expr_mat) <- "numeric"

# ---- clinical ----
cl_patient <- read_cbio_clinical(clp_fp)
cl_sample  <- read_cbio_clinical(cls_fp)

pat_candidates  <- c("PATIENT_ID","Patient.ID","patient_id","PATIENT","PATIENT.ID","PATIENT_IDENTIFIER")
samp_candidates <- c("SAMPLE_ID","Sample.ID","sample_id","SAMPLE","SAMPLE.ID","Tumor_Sample_Barcode","TUMOR_SAMPLE_BARCODE")

pat_col_p  <- pick_col(cl_patient, pat_candidates)
pat_col_s  <- pick_col(cl_sample,  pat_candidates)
samp_col_s <- pick_col(cl_sample,  samp_candidates)

if (is.na(pat_col_p))  pat_col_p  <- pick_by_regex(names(cl_patient), c("patient.*id","^patient","patient"))
if (is.na(pat_col_s))  pat_col_s  <- pick_by_regex(names(cl_sample),  c("patient.*id","^patient","patient"))
if (is.na(samp_col_s)) samp_col_s <- pick_by_regex(names(cl_sample),  c("sample.*id","^sample","barcode","tumor.*sample"))

stopifnot(!is.na(pat_col_p), !is.na(pat_col_s), !is.na(samp_col_s))

cl_patient <- cl_patient %>% dplyr::mutate(PATIENT_ID = as.character(.data[[pat_col_p]]))
cl_sample  <- cl_sample  %>% dplyr::mutate(
  PATIENT_ID = as.character(.data[[pat_col_s]]),
  SAMPLE_ID  = as.character(.data[[samp_col_s]])
)

# ---- match expression cols to SAMPLE_ID ----
expr_samples <- colnames(expr_mat)
expr_samples_norm <- norm_id(expr_samples)
cl_sample <- cl_sample %>% dplyr::mutate(SAMPLE_ID_NORM = norm_id(SAMPLE_ID))

m <- match(expr_samples_norm, cl_sample$SAMPLE_ID_NORM)
matched <- !is.na(m)

expr_use <- expr_mat[, matched, drop = FALSE]
colnames(expr_use) <- cl_sample$SAMPLE_ID[m[matched]]

pheno <- cl_sample[m[matched], , drop = FALSE] %>%
  dplyr::select(SAMPLE_ID, PATIENT_ID, dplyr::everything()) %>%
  dplyr::left_join(cl_patient, by = "PATIENT_ID")

cat("Expression samples (raw):", ncol(expr_mat), "\n")
cat("Matched samples:", ncol(expr_use), " (", round(100*mean(matched),1), "%)\n")

stopifnot(ncol(expr_use) == nrow(pheno))
stopifnot(all(colnames(expr_use) == pheno$SAMPLE_ID))

# Keep in memory for next chunks:
# expr_use, pheno


```




## Chunk 05 — Detect PAM50 column + define labels (PAM50-only pipeline)
This creates:
- `label_pam50_5class` = LumA/LumB/Basal/Her2/Normal (others dropped)
- `label_group2` = ERpos (LumA+LumB) vs TNBC (Basal)  **(optional, but useful)**

## Chunk 05 — Define PAM50 labels (NO GUESSING; explicit nc->NA)
```{r 05_define_PAM50_labels_NO_GUESSING, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(readr)
})

# ----------------------------
# Strict TSV input
# ----------------------------
clin_tsv <- file.path(paths$raw, "brca_metabric_clinical_data.tsv")
stopifnot(file.exists(clin_tsv))

clin_ext <- readr::read_tsv(clin_tsv, show_col_types = FALSE, progress = FALSE)
names(clin_ext) <- make.names(names(clin_ext))

REQ <- c("Sample.ID","Patient.ID","ER.Status","PR.Status","HER2.Status","Pam50...Claudin.low.subtype")
missing_req <- setdiff(REQ, names(clin_ext))
if (length(missing_req) > 0) {
  stop(
    "Missing required columns in brca_metabric_clinical_data.tsv: ",
    paste(missing_req, collapse = ", "),
    "\nAvailable columns:\n", paste(names(clin_ext), collapse = ", ")
  )
}

clin_ext2 <- clin_ext %>%
  transmute(
    SAMPLE_ID_ext   = as.character(.data[["Sample.ID"]]),
    PATIENT_ID_ext  = as.character(.data[["Patient.ID"]]),
    ER_Status_TSV   = as.character(.data[["ER.Status"]]),
    PR_Status_TSV   = as.character(.data[["PR.Status"]]),
    HER2_Status_TSV = as.character(.data[["HER2.Status"]]),
    PAM50_raw       = as.character(.data[["Pam50...Claudin.low.subtype"]])
  )

# ----------------------------
# Strict merge to Chunk 04 objects
# ----------------------------
stopifnot(exists("pheno"), exists("expr_use"))
stopifnot("SAMPLE_ID" %in% names(pheno))
stopifnot(all(colnames(expr_use) == pheno$SAMPLE_ID))

pheno3 <- pheno %>%
  mutate(SAMPLE_ID = as.character(SAMPLE_ID)) %>%
  left_join(clin_ext2, by = c("SAMPLE_ID" = "SAMPLE_ID_ext"))

# Coverage QC
pam_nonempty <- !is.na(pheno3$PAM50_raw) & nzchar(trimws(pheno3$PAM50_raw))
n_pam <- sum(pam_nonempty)
message("PAM50 coverage after merge: ", n_pam, "/", nrow(pheno3),
        " (", round(100*n_pam/nrow(pheno3), 2), "%)")
if (n_pam == 0) stop("Merge failed: 0 samples got PAM50. Check IDs: pheno$SAMPLE_ID vs TSV Sample.ID.")

# ----------------------------
# Strict label validation (NO GUESSING)
# ----------------------------
pam_raw_clean <- tolower(trimws(as.character(pheno3$PAM50_raw)))
pam_raw_clean[pam_raw_clean == ""] <- NA_character_

allowed_raw <- c("luma","lumb","basal","her2","normal","claudin-low","nc")
unexpected <- sort(unique(pam_raw_clean[!is.na(pam_raw_clean) & !(pam_raw_clean %in% allowed_raw)]))
if (length(unexpected) > 0) {
  stop("Unexpected PAM50 labels found:\n- ",
       paste(unexpected, collapse = "\n- "),
       "\n\nFix in TSV or explicitly extend allowed_raw.")
}

# Explicit rule: nc = Not classified => NA
pam_raw_clean[pam_raw_clean == "nc"] <- NA_character_

# Deterministic mapping
pam5 <- case_when(
  pam_raw_clean == "luma"   ~ "LumA",
  pam_raw_clean == "lumb"   ~ "LumB",
  pam_raw_clean == "basal"  ~ "Basal",
  pam_raw_clean == "her2"   ~ "Her2",
  pam_raw_clean == "normal" ~ "Normal",
  TRUE ~ NA_character_          # claudin-low and NA dropped in 5-class
)

pam6 <- case_when(
  pam_raw_clean == "luma"        ~ "LumA",
  pam_raw_clean == "lumb"        ~ "LumB",
  pam_raw_clean == "basal"       ~ "Basal",
  pam_raw_clean == "claudin-low" ~ "Claudin-low",
  pam_raw_clean == "her2"        ~ "Her2",
  pam_raw_clean == "normal"      ~ "Normal",
  TRUE ~ NA_character_
)

# Binary task from PAM6 (Her2/Normal intentionally NA)
group2 <- case_when(
  pam6 %in% c("LumA","LumB")         ~ "ERpos",
  pam6 %in% c("Basal","Claudin-low") ~ "TNBC",
  TRUE ~ NA_character_
)

pheno2 <- pheno3 %>%
  mutate(
    PAM50_col_source   = "brca_metabric_clinical_data.tsv::Pam50...Claudin.low.subtype",
    label_pam50_5class  = factor(pam5, levels = c("LumA","LumB","Basal","Her2","Normal")),
    label_pam50_6class  = factor(pam6, levels = c("LumA","LumB","Basal","Claudin-low","Her2","Normal")),
    label_group2        = factor(group2, levels = c("ERpos","TNBC"))
  )

cat("\nPAM50 raw counts (post nc->NA):\n")
print(sort(table(pam_raw_clean), decreasing = TRUE))

cat("\nPAM50 6-class counts:\n")
print(pheno2 %>% count(label_pam50_6class, name="n") %>% arrange(desc(n)))

cat("\nGroup2 counts:\n")
print(pheno2 %>% count(label_group2, name="n") %>% arrange(desc(n)))

## Chunk 05b — Cohort flow (CONSORT-like) + export

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tibble)
})

stopifnot(exists("expr_mat"), exists("expr_use"), exists("pheno"), exists("pheno2"))
stopifnot(dir.exists(paths$out))

# 1) total expression samples (raw)
n_expr_total <- ncol(expr_mat)

# 2) matched clinical (expression samples that matched cl_sample)
n_matched_clin <- ncol(expr_use)   # equals nrow(pheno)

# 3) PAM50 available (after merge with brca_metabric_clinical_data.tsv)
pam_avail <- !is.na(pheno2$PAM50_raw) & nzchar(trimws(as.character(pheno2$PAM50_raw)))
n_pam_avail <- sum(pam_avail)

# Use PAM50 6-class label for downstream (it keeps Claudin-low + Normal)
y6 <- pheno2$label_pam50_6class

# 4) tumor-only (exclude Normal)
tumor_mask <- !is.na(y6) & (as.character(y6) != "Normal")
n_tumor_only <- sum(tumor_mask)

# 5) 4-class subset (Basal/Her2/LumA/LumB) within tumor-only
primary4 <- c("Basal","Her2","LumA","LumB")
mask4 <- tumor_mask & (as.character(y6) %in% primary4)
n_4class <- sum(mask4)

# Optional extra lines (recommended for transparency)
n_claudinlow <- sum(!is.na(y6) & as.character(y6) == "Claudin-low")
n_normal     <- sum(!is.na(y6) & as.character(y6) == "Normal")
n_nc_or_na   <- nrow(pheno2) - n_pam_avail  # nc mapped to NA + missing

flow <- tibble::tibble(
  step = c(
    "Total samples in expression matrix (raw)",
    "Matched to clinical sample table",
    "PAM50 available after merge (nc -> NA excluded)",
    "Tumor-only subset (exclude Normal)",
    "Primary 4-class subset (Basal/Her2/LumA/LumB)",
    "Tumor-only Claudin-low (secondary/appendix)",
    "Normal samples (excluded from tumor-only)",
    "No PAM50 / not classified (excluded)"
  ),
  n = c(
    n_expr_total,
    n_matched_clin,
    n_pam_avail,
    n_tumor_only,
    n_4class,
    n_claudinlow,
    n_normal,
    n_nc_or_na
  )
)

flow_fp <- file.path(paths$out, "cohort_flow_table.tsv")
readr::write_tsv(flow, flow_fp)

print(flow)
cat("\n✅ Saved cohort flow table:", flow_fp, "\n")


```

```{r 06_task_PAM50_6class_export_for_colab, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tibble)
})

# ------------------------------------------------------------
# Preconditions
# ------------------------------------------------------------
stopifnot(exists("pheno2"), exists("expr_use"), exists("paths"))
stopifnot("SAMPLE_ID" %in% names(pheno2))
stopifnot(all(colnames(expr_use) == pheno2$SAMPLE_ID))
stopifnot(dir.exists(paths$out))

# ------------------------------------------------------------
# TASK: 6-class PAM50 (KEEP Claudin-low)
# Classes: LumA, LumB, Basal, Claudin-low, Her2, Normal
# Drop only NA labels (i.e., nc->NA already in Chunk 05)
# ------------------------------------------------------------
TASK <- "pam50_6class_KEEP_CLAUDINLOW"
label_col <- "label_pam50_6class"
stopifnot(label_col %in% names(pheno2))

keep <- which(!is.na(pheno2[[label_col]]))
ph_final   <- pheno2[keep, , drop = FALSE]
expr_final <- expr_use[, keep, drop = FALSE]
y_final    <- droplevels(ph_final[[label_col]])

stopifnot(ncol(expr_final) == nrow(ph_final))
stopifnot(all(colnames(expr_final) == ph_final$SAMPLE_ID))
stopifnot(length(y_final) == nrow(ph_final))

cat("\nTASK:", TASK, "\n")
cat("Kept samples:", nrow(ph_final), "\n")
cat("Genes:", nrow(expr_final), "\n\n")
print(as.data.frame(table(y_final)))

# ------------------------------------------------------------
# Export formats (READY FOR ML)
#   X_samples_x_genes.tsv.gz : rows = samples, cols = genes
#   y_labels.tsv             : SAMPLE_ID, y
#   meta.tsv                 : SAMPLE_ID, patient + clinical + labels
# ------------------------------------------------------------
X <- t(expr_final)  # samples x genes
stopifnot(nrow(X) == nrow(ph_final))
stopifnot(rownames(X) == ph_final$SAMPLE_ID)

# File paths (exact names you requested)
x_fp    <- file.path(paths$out, "X_samples_x_genes.tsv.gz")
y_fp    <- file.path(paths$out, "y_labels.tsv")
meta_fp <- file.path(paths$out, "meta.tsv")

# X: add SAMPLE_ID as first column; gzip compressed
X_df <- as.data.frame(X, check.names = FALSE) %>%
  tibble::rownames_to_column("SAMPLE_ID")

readr::write_tsv(X_df, file = gzfile(x_fp))

# y: SAMPLE_ID + label string
y_df <- tibble::tibble(
  SAMPLE_ID = ph_final$SAMPLE_ID,
  y = as.character(y_final)
)
readr::write_tsv(y_df, y_fp)

# meta: keep key fields + labels (do NOT guess; only include if present)
meta_df <- ph_final %>%
  transmute(
    SAMPLE_ID,
    PATIENT_ID,
    PAM50_raw,
    label_pam50_6class,
    label_pam50_5class,
    label_group2,
    ER_Status_TSV,
    PR_Status_TSV,
    HER2_Status_TSV
  )

readr::write_tsv(meta_df, meta_fp)

cat("\nSaved exports:\n")
cat("- ", x_fp, "\n", sep = "")
cat("- ", y_fp, "\n", sep = "")
cat("- ", meta_fp, "\n", sep = "")


```


```{r create_release_tree, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Create final reproducibility release tree
# ------------------------------------------------------------

stopifnot(exists("project"), exists("paths"))

release_root <- file.path(project$root, "METABRIC_PAM50_Benchmark_release_v1")

release_dirs <- c(
  "code",
  "code/R",
  "code/python",
  "code/environment",
  "data_processed",
  "results_main",
  "results_secondary",
  "figures_main",
  "figures_supplementary",
  "docs"
)

dir.create(release_root, recursive = TRUE, showWarnings = FALSE)

for (d in release_dirs) {
  dir.create(file.path(release_root, d), recursive = TRUE, showWarnings = FALSE)
}

cat("✅ Release root created:\n", normalizePath(release_root, winslash = "/", mustWork = TRUE), "\n\n")

cat("Subfolders:\n")
print(list.dirs(release_root, recursive = TRUE, full.names = FALSE))

```
```{r sessionInfo_R.txt}
release_root <- "E:/Breast_cancer_DL/METABRIC_PAM50_Benchmark_release_v1"
env_dir <- file.path(release_root, "code", "environment")
dir.create(env_dir, recursive = TRUE, showWarnings = FALSE)

writeLines(capture.output(sessionInfo()),
           file.path(env_dir, "sessionInfo_R.txt"))

pkgs <- as.data.frame(installed.packages()[, c("Package", "Version")])
pkgs <- pkgs[order(pkgs$Package), ]
readr::write_tsv(pkgs, file.path(env_dir, "installed_packages_R.tsv"))

cat("Saved:\n")
cat("-", file.path(env_dir, "sessionInfo_R.txt"), "\n")
cat("-", file.path(env_dir, "installed_packages_R.tsv"), "\n")

```



